Extending statistical models for source attribution of zoonotic diseases: A study of campylobacteriosis

### 2018-03-19 paper/finalish/article_1023.tex
updates discussion

\section{Discussions}
Models that determine the source of human infection, particularly with zoonotic pathogens that originate in animal populations, are of considerable value to public health policy makers. However, the complexity of models may increase due to the complex transmission pathways of pathogens, particularly when a model combines epidemiological and molecular data. An outstanding question is whether such complexity is required, or whether a simpler model may work as effectively. Here we developed a relatively simple model to estimate the attribution probability for each source of \textit{Campylobacter} infection. This model differs to the asymmetric Island model, in that it does not model pathogenic evolution, opting instead to use infer the sampling distribution of genotypes directly from the observed count data.

Our results showed that the simple Dirichlet model and the asymmetric Island model showed largely similar results for the final attribution, with both models demonstrating a clear effect of rurality on attribution: cases in rural areas are more likely to have originated from ruminants, while those in main urban centers are more likely to be of poultry origin. While the overall attribution was consistent, it would be expected that the conditional probabilities for a given genotype might differ markedly between models. For those genotypes observed infrequently (or not at all) among the sources, the Dirichlet model has little information, while the asymmeteric Island model can exploit additional information from cases with similar (but not identical) genetic profiles. In the case of MLST data, with just 7 loci, the majority of human cases and source isolates come from a relatively small number of sequence types which are observed often. Thus the Dirichlet model performs well, as it has sufficient observations to estimate the genotype distribution well where the bulk of the data lie. It is only those genotypes that are rarely observed where it performs poorly, but as they are rarely observed, they don't contribute much to the overall attribution. In other circumstances, such as where we have many more than 7 loci, we would have many more rare genotypes, so that the Dirichlet model might provide little useful information. At the extreme example of whole genome MLST (wgMLST) where each isolate might be unique, it would provide essentially no information at all. In these circumstances, however, the asymmetric Island model would be expected to still perform well, assuming that information could be transferred between similar genotypes.

As with any Bayesian analysis, it is of interest to examine the sensitivity of hte results to the choice of prior distributions. We originally used standard normal priors for regression coefficients on the logit scale. We also considered...

In light of the different estimates of the sampling distribution of genotypes from the asymmetric Island and the simple models, two approaches of model fitting (linear and non-linear) were utilized to estimate the attribution probability via the Metropolis-Hastings algorithm. Under the assumption of a logit scale on the attribution probability, we examined how the values of $f$ affected the resulting attribution probability. We changed the standard deviation of the prior distribution from 1 to 2 in the algorithm. The associated posterior mean of attribution probability for each source was invariant despite the fact that the proposals of $f$ jumped even wider in the random walk. This is because the denominator of the equation~\eqref{capF} balances the impact made by the values of $f$. Further, we included a rurality covariate in the model fitting. Our findings not only coincided with the hypothesis that poultry is the major cause of human campylobacteriosis in urban areas \cite{MullA, Marsh, MullM, Leve} but also revealed the linear geographic effect on the infection. It means that poultry is less likely responsible for the infection when cases dwell away from urban areas. Furthermore, the result of decreasing poultry contribution in main urban areas from fitting the data with post-intervention period accorded with what Sears A. and other researchers investigated the effect of poultry intervention \cite{AnnS}. This could provide additional information that would help policy-makers to target interventions to populations residing in urban compared to rural areas.

In conclusion, the simple model is comparable to the asymmetric Island model. The potential of the inclusion of covariates in the model could be applied to help make a decision on the control of the infection. In addition, the simple model (although based on the genetic data) does not capture the evolutionary processes that could be also applied to any type of classification such as phenotype in the research of interest. Nonetheless, there is some concern to apply our approach. One is the role of water that has been recently addressed in the literature \cite{Wagen} by treating it as a medium instead in the transmission network. The other is the larger pathogen datasets resulted from rapidly evolving molecular tools such as the extending MLST schemes typing from seven loci to whole-genome MLST. Hence, reconsidering the role of water as a source of infection and selecting the invaluable information from such large pathogenic data to fit the extended model will be the potential challenges in future research.

>>>>

\section{Discussion}
Models that determine the source of human infection, particularly with zoonotic pathogens that originate in animal populations, are of considerable value to public health policy makers. However, the complexity of models may increase due to the complex transmission pathways of pathogens, particularly when a model combines epidemiological and molecular data. An outstanding question is whether such complexity is required, or whether a simpler model may work as effectively. Here we developed a relatively simple model to estimate the attribution probability for each source of \textit{Campylobacter} infection. This model differs to the asymmetric Island model, in that it does not model pathogenic evolution, opting instead to use infer the sampling distribution of genotypes directly from the observed count data.

Our results showed that the simple Dirichlet model and the asymmetric Island model showed largely similar results for the final attribution, with both models demonstrating a clear effect of rurality on attribution: cases in rural areas are more likely to have originated from ruminants, while those in main urban centers are more likely to be of poultry origin. As most people in the Manawatu region live in urban centers, this highlights the importance of poultry as a reservoir for campylobacteriosis, which is well established in the literature \cite{MullA, Marsh, MullM, Leve}.

When we ran models allowing attribution probabilities to differ before and after the intervention in the poultry industry in 2007, we saw a clear difference, with much lower poultry (and higher ruminant) attribution, particularly in main urban centers. Considering that most people in the Manawatu region live in urban areas, and that case rates in urban areas decreased from 2008 onwards, it is clear that this intervention coincided with a dramatic reduction in poultry attributed illness as reported elsewhere \cite{AnnS}. Given that campylobacteriosis cases in rural areas are mostly attributed to ruminant sources, and that case rates in these areas have been higher than those in urban centers since 2008, there is a clear need for public health interventions to focus in this area.

While the overall attribution was consistent between the Dirichlet and asymmetric Island models, it would be expected that the conditional probabilities for a given genotype might differ markedly. For those genotypes observed infrequently (or not at all) among the sources, the Dirichlet model has little information while the asymmeteric Island model can exploit information from cases with similar (but not identical) genetic profiles. In the case of MLST data with just 7 loci, the majority of human cases and source isolates come from a relatively small number of sequence types which are observed often. Thus the Dirichlet model performs well, as it has sufficient observations to estimate the genotype distribution where the bulk of the data lie. It is only those genotypes that are rarely observed where it performs poorly, but as they are rarely observed, they don't contribute significantly to the overall attribution. In other circumstances, such as where we have many more than 7 loci, we would expect to have many more rare genotypes, so that the Dirichlet model might provide little useful information. At the extreme example of whole genome MLST (wgMLST) where each isolate would typically be unique, it would provide essentially no information at all. In such circumstances, however, the asymmetric Island model would be expected to still perform well, assuming that information could still be transferred between similar genotypes.

As with any Bayesian analysis, it is of interest to examine the sensitivity of the results to the choice of prior distributions. We originally used standard normal priors for regression coefficients on the logit scale. We also considered priors with $\sigma^2=4$, and while this meant that $f$ tended to drift further from 0, the resulting attribution probabilities $F$ did not, largely as the attribution is dominated by poultry and ruminant sources, with the water source in particular being close to zero. Thus, $f_\mathsf{poultry}$ and $f_\mathsf{ruminants}$ are positive, while $f_\mathsf{water}$ is negative and the magnitude of these can increase without making all that much difference to their corresponding $F$'s. The prior on $f$ thus tends to restrict this ill-behaviour rather than acting as a strong constraint on attribution probabilities. The prior $\mathbf{\alpha}_j$ in the Dirichlet also makes little difference if kept small, as it most strongly effects genotypes that are rare, which do not contribute significantly to the overall attribution. It can be thought of as augmenting with new data, so that $\mathbf{\alpha}_j=\mathbf{1}$ is equivalent to adding a single observation of each genotype to source $j$. Thus, large values of $\mathbf{\alpha}_j$ will cause the genotype distributions across sources to look more similar, and hence result in equal attribution to each source.

Future research might focus on possible extensions to these models. One direction is to adapt the models further with additional covariates, which might include age, occupation, and other risk factors such as contact with animals. For example, there is evidence that children in rural areas are at higher risk of campylobacteriosis through contact with farm animals \cite{MullM, Spencer2012}. Another direction is in expanding the role of water in particular. In these models, we've assumed that water is a source of human campylobacteriosis infection, but water differs from the other food and environmental sources in that it is not an amplifying reservoir for \emph{Campylobacter} \cite{Wagen}. In contrast, the genotypes found in water might be expected to originate in the other sources present here, particularly ruminants and wild birds, but also potentially from humans as well via discharge of unprocessed human waste. Hence, water acts as a transmission pathway from sources to humans, being both an end point (reduced water quality from faecal contamination) and a source (human consumption of water, either recreationally or through untreated water supplies). While there is presently little evidence that water is an important source for human campylobacteriosis from the current models, these are fit on sporadic cases of campylobacteriosis. However, water is known as a key source of outbreaks of campylobacteriosis, such as the large outbreak in Havelock North, New Zealand in 2016 where an estimated 5,500 out of 14,000 residents became ill \cite{dia_campy_report1}. Thus, characterising both characterising the source of \emph{Campylobacter} found in water has important implications for both water quality and public health.
